---
title: "A survey of ICD9 term usage in the MIMIC-III EHR Data set"
output: html_notebook
---

Overall research question: What are the characteristics of ICD9 code usage, and are some conceptual reasons that help explain that usage?
  - What are some characteristics to explore
    - the count of each term - break terms down into low, medium and high term
    - term usage by different groups of terms within the hierarchy
    - term usage by different patient populations
  - Conceptual reasons
    - term granualarity 
    - financial biases 


## ICD9 Term counts 

```{r}
library(tidyverse)
load('Rdata/icd_code_reference_and_counts.Rdata')
```


There are 12167 total ICD9 terms. The number of them that are used in mimic:

```{r}
sum(icd_code_counts_annotated$count != 0)
```

So a little over half the ICD9 terms are not used.

This is what the distribution of counts looks like for the unused terms 

```{r}
icd_code_counts_annotated %>% filter(count!=0) %>% 
  mutate(k='ICD9') %>% 
ggplot(aes(x=k, y=count))+
  geom_violin()+
  geom_boxplot(width=.05)+
  xlab('')+
  ylim(c(0, 100))+
  theme_classic() +
  theme(text = element_text(size = 15))


```

Some terms are not likely not used because the disease that it describe is not likley to be seen in the hospital. Ideally, I would like to remove these form further analysis, but identifying them is a difficult task. 

The stategy I will adopt is to filter based on usage of parent codes:
Each ICD9 consists of a 3 digit parent code, and 2 digit subcode; Each parent code is supposed to represent a broad class of disease. I'm going going to keep a term if its associated parent code has at least one used term across all its children. 

```{r}
icd_code_counts_annotated %>% group_by(parent_code) %>% 
  summarise(parent_never_used = sum(count) == 0) %>% filter(parent_never_used) %>% nrow
icd_code_counts_annotated %>% group_by(parent_code) %>% 
  summarise(parent_never_used = sum(count) == 0) %>% filter(parent_never_used) %>% 
  inner_join(icd_code_counts_annotated) %>% nrow

codes_to_remove <- icd_code_counts_annotated %>% group_by(parent_code) %>% 
  summarise(parent_never_used = sum(count) == 0) %>% filter(parent_never_used) %>% pull(parent_code)
```
So 83 parent codes out of the 928 total have no usage across all their children.

These 83 parent codes contain 633 total ICD code that will be excluded

### depth of term in the snomed hierarchy

We can map terms to snomed hierarchy, and then calculate the distance of each term to the root node. We can use this as a proxy for granularity.


## ICD 2 SNOMED

```{r}
load('Rdata/ICD9_ref_mapped_to_snomed.Rdata')
snomed_term_depth = read_csv('SNOMED_term_depth_from_networkx.csv', col_types = 'cd', col_names = c("SNOMED_CID", "depth"), skip=1)
ICD9_code_depth <- inner_join(icd_reference_with_snomed, snomed_term_depth) %>% group_by(ICD9_CODE) %>% summarise(avg_depth = mean(depth))

```




```{r}
icd_counts_annotated_with_depth <- icd_code_counts_annotated %>% 
  filter(!parent_code %in% codes_to_remove) %>% 
  mutate(is_unused = ifelse(count == 0, 'Unused', 'Used')) %>% 
  inner_join(ICD9_code_depth)

sum(is.na(icd_counts_annotated_with_depth$avg_depth))
```

so roughly  1700 terms do have a mapping to snomed

```{r}
icd_counts_annotated_with_depth <- icd_counts_annotated_with_depth %>% 
  filter(!is.na(avg_depth))
ggplot(icd_counts_annotated_with_depth) +
  geom_boxplot(aes(x=is_unused, y=avg_depth))
```

So at first glance, it seems like there's no difference between the granularity of used and un-used terms 




However, if we split terms up by their broader disease category, we can see that different classes of terms lead to different depths.

```{r fig.height=7}
ggplot(icd_counts_annotated_with_depth) +
  geom_boxplot(aes(x=parent_code_class, y=avg_depth)) + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

This leads me to hypothesize that there's relative scaling factor specific to disease

So instead, I'm going to compare depth only within a parent code

```{r}
depth_by_pcode_and_usage <- icd_counts_annotated_with_depth %>% group_by(parent_code, is_used) %>% 
  summarise(med_depth = median(avg_depth)) %>% 
  pivot_wider(names_from = is_used, values_from = med_depth) %>% 
  filter( !(is.na(Unused)|is.na(Used)) ) %>% 
  mutate(delta = Unused - Used) 

summary(depth_by_pcode_and_usage$delta)

class2parent <- icd_code_counts_annotated %>% select(parent_code, parent_code_class) %>% distinct

depth_by_pcode_and_usage %>% 
  mutate(k='df', 
         delta = Unused-Used) %>% 
  inner_join(class2parent) %>% 
ggplot() + 
    geom_violin(aes(x=k, y=delta), width = .5)+
    geom_boxplot(aes(x=k, y=delta), width = .05) +
    geom_point(data = tibble(k="df", y= mean(depth_by_pcode_and_usage$delta)),aes(x=k, y=y), color = "red")+
geom_hline(yintercept = 0)




```


This still isnt super meaningful, so lets try one more thing. Perhaps there might be a granularity difference between terms that *are* used 

Now lets try and better differentiate between used - split it into bins 

bins
```{r}
sum(icd_code_counts_annotated$count == 1)
sum(icd_code_counts_annotated$count >=2 & icd_code_counts_annotated$count <=5)
 
sum(icd_code_counts_annotated$count >5 &icd_code_counts_annotated$count <= 25)

sum(icd_code_counts_annotated$count > 25)
```

```{r}

tdiff <- function(x){
  if(length(x) >1){
   combn(x, 2) %>% t %>% {paste(.[,1] , .[,2], sep = " - " )}
  }else{
    0
  }
  
}

tcompn <- function(x){
  if(length(x) >1){
   combn(x, 2) %>% t %>% {paste(.[,1] , .[,2], sep = " - " )}
  }else{
    paste(x, sep = " - ")
  }
}

depth_by_bin <-  icd_counts_annotated_with_depth %>% 
  filter(!is.na(avg_depth)) %>% 
  mutate(count_bin = case_when(
    count ==  0 ~ "0",
    count == 1 ~ "1", 
    count >=2 & count <5 ~ "[2,5)",
    count >=5 & count <25 ~ "[5, 25)",
    count >= 25 ~ ">=25"
  ) ) 
ggplot(depth_by_bin)+
  geom_boxplot(aes(x = count_bin, y= avg_depth)) + 
  theme_minimal()

ggplot(icd_counts_annotated_with_depth %>% filter(), aes(x= count, y = avg_depth)) + 
  geom_point() + 
  geom_smooth(method = 'lm') + 
  xlim(c(1,100))
depth_by_bin %>% 
  group_by(parent_code, count_bin) %>% 
  summarise(med_depth = median(avg_depth)) %>% 
  group_by(parent_code) %>% 
  summarise(dm = c(dist(med_depth, method = "manhattan")), 
            comp = tcompn(count_bin))
  # pivot_wider(names_from = is_used, values_from = med_depth) %>% 
  # filter( !(is.na(Unused)|is.na(Used)) ) %>% 
  # mutate(delta = Unused - Used) 


sum(is.na(icd_counts_annotated_with_depth$avg_depth))

c(dist(c('a' = 0, 'b' = 2, 'c' = 3 ,'d' = -2 )))

```

