---
title: "R Notebook"
output: html_notebook
---


https://www.cms.gov/Medicare/Medicare-Fee-for-Service-Payment/AcuteInpatientPPS/MS-DRG-Classifications-and-Software
this gets you the ICD10 > DRG mapping 


https://www.cms.gov/Medicare/Medicare-Fee-for-Service-Payment/AcuteInpatientPPS/Downloads/FY2020-FR-Table-5.zip 
This gets you DRG weights 

medicare claims data 
https://data.cms.gov/provider-summary-by-type-of-service/medicare-inpatient-hospitals/medicare-inpatient-hospitals-by-provider-and-service/data/2014


Okay figured out why the DRG descriptions dont match what we have found : THERE ARE TWO(actually 3) KINDS OF DRG CODES!!!!!

CMS/HCFA DRG: older standard used < 2007/8
MS-DRG: current codes
Same 3 digit system, but completely different meanings. Whoever did this should be fired

APR DRG: uses same codes but adds extra stuff for more granularity; these are *in addition* to the other two; ignore these for now

```{r}
library(tidyverse)
library(DBI)

con = dbConnect(RSQLite::SQLite(), 'data/mimic3-1.4/mimic3.db')
con %>% tbl('DRGCODES') %>% collect
tables = read_csv('data/mimic3-1.4/tbl_headers.csv') %>% .[[1]]
load('Rdata/icd_code_reference_and_counts.Rdata')
DRG_codes_raw <- con %>% 
  tbl('DRGCODES') %>% 
  select(-ROW_ID) %>%  
  collect %>% # DRG codes are only supposed to be 3 digits, the 4 digit ones have the severity stuck on the end  
  mutate(DRG_CODE = substr(DRG_CODE, 1, 3))


DRG_codes_raw %>% select(SUBJECT_ID, HADM_ID) %>% distinct %>% nrow


table(DRG_codes_raw$DRG_TYPE)
DRG_codes_raw %>% filter(DRG_TYPE != "APR") %>% select(SUBJECT_ID, HADM_ID) %>% distinct %>% nrow

```

```{r}
drg_tab <-  DRG_codes_raw %>% select(DRG_TYPE, DRG_CODE) %>% distinct() %>% group_by(DRG_TYPE) %>% 
  summarise(codeset = list(unique(DRG_CODE)))
N_d_HCFA = drg_tab[2, 2] %>% .[[1]] %>% .[[1]] %>%  length
N_d_MS = drg_tab[3, 2] %>% .[[1]] %>% .[[1]] %>%  length
```


Map CMS-DRG to MS-DRG

found mapping here https://www.kmap-state-ks.us/documents/content/provider/ms-drg%20crosswalk.pdf

```{r}
raw_cms_ms_mapping <- read_csv('~/Downloads/tabula-cms-drg_to_ms-drg.csv', col_names = F) %>% 
  filter(! rowSums(is.na(.)) == ncol(.)   ) %>% 
  mutate(rowid = 1:nrow(.))

mapping_off_1col <- raw_cms_ms_mapping[c(659:677, 724:794), ]


HCFA_to_MS_drg <- bind_rows(
    raw_cms_ms_mapping %>% filter(!is.na(X1) & !is.na(X4)) %>% select(CMS=X1, MS=X4),
  mapping_off_1col %>% filter(!is.na(X1) & !is.na(X5)) %>% select(CMS=X1, MS=X5)
) %>% 
  mutate(CMS = as.numeric(CMS), 
         MS = as.numeric(MS)) %>% 
  filter(!is.na(CMS) & !is.na(MS)) %>% 
  mutate(HCFA = paste0('HC', CMS), 
         MS = paste0("MS", MS)) %>% 
  select(-CMS) %>% 
  distinct()


sum(paste0("MS", drg_tab$codeset[[3]]) %in% HCFA_to_MS_drg$MS)
N_hcfa_mapped <- sum(paste0("HC", drg_tab$codeset[[3]]) %in% HCFA_to_MS_drg$HCFA)

```

Clean up DRG table 

```{r}
DRG_codes_HCFA <- DRG_codes_raw %>% filter(DRG_TYPE == "HCFA") %>% 
  mutate(HCFA =paste0('HC', DRG_CODE) ) %>% 
  left_join(HCFA_to_MS_drg) %>% 
  select(-DRG_CODE, -HCFA) %>% 
  rename(DRG_CODE = MS)
sum(is.na(DRG_codes_HCFA$DRG_CODE))

diag_DRG_cleaned <-  bind_rows(
  DRG_codes_raw %>% filter(DRG_TYPE == "MS") %>% 
    mutate(DRG_CODE = paste0("MS", DRG_CODE)),
  DRG_codes_HCFA %>% filter(!is.na(DRG_CODE))
) %>% rename(DRG_DESCR = DESCRIPTION)
diag_DRG_cleaned %>% select(SUBJECT_ID, HADM_ID) %>% distinct %>% nrow
```



Now lets tie these to dollar amounts

MIMIC ICD 9  to DRG mapping

```{r}
load('Rdata/ICD9_diag_codes_by_patient_visit.Rdata')

diag_ICD_DRG <- diag_with_admits %>% 
  inner_join(diag_DRG_cleaned %>% select(-DRG_SEVERITY, -DRG_MORTALITY))


```



Method 1: Using medicare claims data 


```{r}
all_medicare_billing_data <- list.files(pattern = "Medicare_Inpatient_Hospitals_by_Provider_and_Service_Data") %>% 
  lapply(function(x) read_csv(x, col_types = rep("c", 15) %>% paste(collapse =""))) %>% bind_rows
sum(unique(all_medicare_billing_data$DRG_Cd) %in% unique(DRG_codes_raw$DRG_CODE))




unique(DRG_codes_raw$DRG_CODE) %>% length
medicare_avg_cost_per_DRG <- all_medicare_billing_data %>% 
  mutate(Avg_Submtd_Cvrd_Chrg = as.numeric(Avg_Submtd_Cvrd_Chrg), 
         Avg_Tot_Pymt_Amt = as.numeric(Avg_Tot_Pymt_Amt),
         Avg_Pymt_Delta = Avg_Tot_Pymt_Amt -  Avg_Submtd_Cvrd_Chrg ) %>% 
  group_by(DRG_Cd) %>% summarise(avg_dollar_sub = mean(Avg_Submtd_Cvrd_Chrg, na.rm =T), 
                                 avg_dollar_rec = mean(Avg_Tot_Pymt_Amt, na.rm = T),
                                 avg_ri_delta = mean(Avg_Pymt_Delta, na.rm = T)) %>% 
  filter(rowSums(is.na(.)) == 0 ) %>% 
  mutate(DRG_CODE = paste0("MS", DRG_Cd)) %>% 
  select(-DRG_Cd)

N_medicare_DRG <-  n_distinct(medicare_avg_cost_per_DRG$DRG_CODE)
```


```{r}
decompose_code_weights <- function(tot, prio_seq){
  prio_seq = as.numeric(prio_seq)
  cw = sum(1/prio_seq)
  x = tot/cw 
  return( x * 1/prio_seq)
}
ICD9_mapped_to_medicare_dollar <-  inner_join(diag_ICD_DRG, medicare_avg_cost_per_DRG) %>% 
  group_by(SUBJECT_ID, HADM_ID, DRG_CODE) %>% 
  summarise(SEQ_NUM = SEQ_NUM, 
            ICD9_CODE = ICD9_CODE, 
            DRG_TYPE = DRG_TYPE, 
            DRG_CODE = DRG_CODE,
            avg_dollar_sub=avg_dollar_sub,
            avg_dollar_rec=avg_dollar_rec,
            avg_ri_delta =avg_ri_delta,
            ICD9_dc_avg_dollar_sub = decompose_code_weights(first(avg_dollar_sub), SEQ_NUM),
            ICD9_dc_avg_dollar_rec = decompose_code_weights(first(avg_dollar_rec), SEQ_NUM),
            ICD9_dc_avg_ri_delta = decompose_code_weights(first(avg_ri_delta), SEQ_NUM)
            ) 

estim_dollar_per_ICD9_medicare <- ICD9_mapped_to_medicare_dollar %>% group_by(ICD9_CODE) %>% 
  summarise(avg_estim_dollar_sub = mean(ICD9_dc_avg_dollar_sub), 
            avg_estim_dollar_rec = mean(ICD9_dc_avg_dollar_rec),
            avg_estim_dollar_ri = mean(ICD9_dc_avg_ri_delta)
            )

N_DRG_mapped_cost <-  ICD9_mapped_to_medicare_dollar$DRG_CODE %>% n_distinct 

N_case_drg_cost <-  diag_DRG_cleaned %>% filter(DRG_CODE %in% ICD9_mapped_to_medicare_dollar$DRG_CODE) %>% select(SUBJECT_ID, HADM_ID) %>% distinct %>% nrow

N_icd_w_cost <-  estim_dollar_per_ICD9_medicare$ICD9_CODE %>% n_distinct

```

```{r}
N_medicare_DRG = all_medicare_billing_data %>% pull(DRG_Cd) %>% n_distinct()
N_mimic_DRG = DRG_codes_raw %>% pull(DRG_CODE) %>% n_distinct()
N_commmon_DRG = length(intersect(all_medicare_billing_data$DRG_Cd, DRG_codes_raw$DRG_CODE))

```

```{r, fig.width =4}
ggplot(estim_dollar_per_ICD9_medicare %>% mutate(x='')) + 
  geom_boxplot(aes(x=x, y=avg_estim_dollar_sub), width=.5) + 
  ylim(c(0, 20000)) + 
  ylab("Average Submitted Cost") +
  xlab("ICD9 Code") +  
  theme_minimal()
  
```

```{r}
ggplot(estim_dollar_per_ICD9_medicare %>% mutate(x='') %>% mutate(frac_ri = avg_estim_dollar_rec / avg_estim_dollar_sub)) + 
  geom_boxplot(aes(x=x, y=frac_ri), width=.5) + 
  ylab("Average Submitted Cost") +
  xlab("ICD9 Code") +   
  theme_minimal()
  
```

```{r}
save( estim_dollar_per_ICD9_medicare, file = "Rdata/ICD9_estimated_dollar_amounts.Rdata")
```


```{r}
estim_dollar_per_ICD9_medicare %>% arrange(desc(avg_estim_dollar_sub)) %>% tail
```


```{r}
all_medicare_billing_data %>% mutate(frac_ri = Avg_Mdcr_Pymt_Amt %>% as.numeric() / Avg_Submtd_Cvrd_Chrg %>% as.numeric(), x='') %>% 
  filter(frac_ri <= 1) %>% 
  ggplot() + 
  geom_boxplot(aes(x=x, y=frac_ri)) + 
  xlab('Medicare DRG Code') + 
  ylab("Fraction Reimbursed")

N_average_medicare_RI = all_medicare_billing_data %>% mutate(frac_ri = Avg_Mdcr_Pymt_Amt %>% as.numeric() / Avg_Submtd_Cvrd_Chrg %>% as.numeric(), x='') %>% 
  filter(frac_ri <= 1) %>% pull(frac_ri) %>% mean(na.rm = T) %>% round(2)
```


```{r}
cost_with_counts <-  inner_join(icd_code_counts_annotated,  estim_dollar_per_ICD9_medicare) %>% 
  mutate(frac_rec = avg_estim_dollar_rec/avg_estim_dollar_sub, 
         j= 1/avg_estim_dollar_sub)

spcor <- cost_with_counts %>% select(count, avg_estim_dollar_sub) %>% cor(method = 'spearman') %>% .[1,2] %>% round(2)

ggplot(cost_with_counts, aes(x=count, y = avg_estim_dollar_sub)) + 
  xlim(c(0, 1000))+
  ylim(c(0,100000))+
  geom_point(alpha = .2)  + 
  geom_label(data = tibble(count =750 , avg_estim_dollar_sub = 75000, text = paste("Spearman correlation:", spcor)),
             aes(x=count, y = avg_estim_dollar_sub, label = text)
             )+
  xlab("ICD9 Code Count") + 
  ylab("Average Submitted Cost")+
  theme_minimal() + 
  theme_minimal()
mpdf = tibble(y=estim_dollar_per_ICD9_medicare %>% pull(avg_estim_dollar_sub ) %>% mean, 
              x='')
N_mean_cost = round(mpdf$y, 2)
```

```{r}
cost_with_counts_binned <-  cost_with_counts %>%   mutate(count_bin = case_when(
    count ==  0 ~ "0",
    count == 1 ~ "1", 
    count >=2 & count <5 ~ "[2,5)",
    count >=5 & count <25 ~ "[5, 25)",
    count >= 25 ~ ">=25"
  ) ) %>% mutate(count_bin = factor(count_bin, levels = c("0","1","[2,5)","[5, 25)",">=25"))) 
avg_cost_per_bin <- cost_with_counts_binned %>% group_by(count_bin) %>% 
  summarise(avg_cost = mean(avg_estim_dollar_sub)) %>% 
  mutate(k="Average Cost")
N_aov_pval=summary(aov(avg_estim_dollar_sub ~ count_bin, data = cost_with_counts_binned ) )[[1]][["Pr(>F)"]][1] %>% round(5)
aov_lab <- tibble(x=1.5, y=22000, text = N_aov_pval %>% {paste0("Anova p value = ",.)} )
ggplot(cost_with_counts_binned) + 
  geom_boxplot(aes(x=count_bin, y=avg_estim_dollar_sub), outlier.shape = NA) + 
  geom_point(data = avg_cost_per_bin, aes(x=count_bin, y=avg_cost ,color =k ))+
  geom_label(data = aov_lab, aes(x=x, y=y, label =text))+
  guides(color = guide_legend(title = ""))+
  theme_minimal()+
  ggtitle("Cost per count bin (outliers not shown)")+
  ylim(c(0,25000)) + 
  xlab("Count Bin") + 
  ylab("Average Submitted Cost")


N_cost_outliers <-  sum(estim_dollar_per_ICD9_medicare$avg_estim_dollar_sub > 25000)
```
```{r}
cost_with_counts_binned %>% filter(count_bin == "1") %>% View
```



```{r}
save(cost_with_counts_binned,aov_lab, avg_cost_per_bin, cost_with_counts, spcor, estim_dollar_per_ICD9_medicare, all_medicare_billing_data, file = 'Rdata/ICD9_and_DRG_exploration_plotdata.Rdata'  )
save(list = ls()[grepl('N_', ls())], file = 'Rdata/ICD9_and_DRG_exploration_countdata.Rdata' )

```





Below here not relevant 


