---
title: "R Notebook"
output: html_notebook
---

### Mimic III background
- paper https://www.nature.com/articles/sdata201635
- only patients from the ICU
- only contains adults (age > 16) and and neonates(age<???)
- sounds like they kept everyone that mwt some wort of data quality threshsold?

## Notes 

-  link for IC9 <==> SNOMED-CT mapping https://www.nlm.nih.gov/research/umls/mapping_projects/icd9cm_to_snomedct.html


## TODO

- Seems like parent level ICD9 codes are never used, unless the parent terms has no direct children 
- Mimic is restricted to just ICU patients > 16, and a sprinkling of neonates. I should remove the terms that are likely not relevant to the ICU
  - one possibility is mapping ICD-9 codes to SNOMED-CT, and seeing if we can infer some sort of pediatric event from there?


Restructured data into sqlite db 

```{r}
library(tidyverse)
library(DBI)

con = dbConnect(RSQLite::SQLite(), 'data/mimic3-1.4/mimic3.db')
con %>% tbl('ADMISSIONS') %>% collect
tables = read_csv('data/mimic3-1.4/tbl_headers.csv') %>% .[[1]]
```

### ICD9 Diagnoses 
ICD9 Diagnoses are in `DIAGNOSES_ICD`; references are in `D_ICD_DIAGNOSES`
```{r}
diag_icd =  con %>% tbl('DIAGNOSES_ICD') %>% select(-ROW_ID) %>%  collect 
icd_reference = con %>% tbl('D_ICD_DIAGNOSES')%>% select(-ROW_ID) %>% collect()
n_distinct(icd_reference$ICD9_CODE) 
```

Not all ICD9 codes are present here. We can pull more from the mappers we will use later 


```{r}


i2s_1t1 <- read_tsv('~/Downloads/ICD9CM_DIAGNOSIS_MAP_202012/ICD9CM_SNOMED_MAP_1TO1_202012.txt', col_types = 'ccccccccccc') %>% mutate(type = '1-to-1')
i2s_1tm <- read_tsv('~/Downloads/ICD9CM_DIAGNOSIS_MAP_202012/ICD9CM_SNOMED_MAP_1TOM_202012.txt', col_types = 'ccccccccccc') %>% mutate(type = '1-to-Many')
icd_to_SNOMED_mapping <- bind_rows(i2s_1tm, i2s_1t1) %>% 
  mutate(ICD9_CODE = str_replace_all(ICD_CODE, '\\.', ''),
         SNOMED_CID = as.character(SNOMED_CID))
icd_reference <- bind_rows(
  icd_reference,
  icd_to_SNOMED_mapping %>% filter(!ICD9_CODE %in% icd_reference$ICD9_CODE) %>% select(ICD9_CODE, SHORT_TITLE=ICD_NAME) %>% distinct
)
```


For some now, I'm going to ignore the E and V supplemental codes 

```{r}
library(lubridate)
admissions = con %>% 
  tbl('ADMISSIONS') %>% 
  select(all_of(colnames(.)[1:5])) %>% collect() %>% select(-ROW_ID)

admissions %>% group_by(SUBJECT_ID, HADM_ID) %>% count ## each row is a distinct individual,admit combo
diag_with_admits <-  admissions %>% left_join(diag_icd) %>% mutate(ADMITTIME = as_datetime(ADMITTIME), 
                                                                   DISCHTIME = as_datetime(DISCHTIME)
                                                                   )
```
rationale about how to use counts 

```{r}

count_strat_validation_data <- bind_rows(
  diag_with_admits %>% group_by(SUBJECT_ID) %>% count %>% mutate(x="# of visits per patient"),
  diag_with_admits %>% group_by(SUBJECT_ID, ICD9_CODE) %>% count %>% mutate(x="# of times ICD9 code billed per patient") 
)

N_frac_total_used_once_pp <-  count_strat_validation_data %>% filter(x=="# of times ICD9 code billed per patient") %>% pull(n) %>% 
  {sum(.[. == 1])/sum(.)} %>% {.*100} %>% round(2)

N_avg_visits_pp <- count_strat_validation_data %>% filter(x!="# of times ICD9 code billed per patient") %>% pull(n) %>% mean  %>%  round(2)
N_cases =diag_with_admits %>% select(SUBJECT_ID, HADM_ID) %>% distinct %>% nrow

```



We'll define a count as each distinct entry of a code; in the future, we might consider distinct per patient 

```{r}

icd_code_counts <- diag_with_admits %>% group_by( ICD9_CODE) %>% count  %>% rename(count =n)
```

majority of terms have less than 7 counts 

```{r}

icd_reference_annotated =icd_reference %>% 
  mutate(icd_numeric = as.numeric(ICD9_CODE), 
         icd_numeric = replace(icd_numeric, grepl("E", ICD9_CODE), -1),
         icd_numeric = replace(icd_numeric, grepl("V", ICD9_CODE), -2),
         is_used = ifelse(ICD9_CODE %in% diag_icd$ICD9_CODE,'Used', 'Unused'),
         code_str = as.character(ICD9_CODE), 
         code_len = nchar(code_str),
         is_parent = ifelse(code_len > 3, 'child', 'parent'),
         parent_code = substr(code_str, 1, 3),
         parent_code_num = as.numeric(parent_code),
         parent_code_num = replace(parent_code_num, is.na(parent_code_num), icd_numeric[is.na(parent_code_num)]),
         parent_code_class =  case_when(
                    parent_code_num == -1 ~ "External Injury",
                    parent_code_num == -2 ~ "Supplemental classification of patient background",
                    parent_code_num>=1 & parent_code_num<=139 ~ 'Infectious and Parasitic disease',
                    parent_code_num>=140 & parent_code_num<=239 ~ 'Neoplasms', 
                    parent_code_num>=240 & parent_code_num<=279 ~ 'Endocrine, Metabolic, Immunity Disorders',
                    parent_code_num>=280 & parent_code_num<=319 ~ 'Blood diseases',
                    parent_code_num>=320 & parent_code_num<=389 ~ 'brain disease',
                    parent_code_num>=390 & parent_code_num<=459 ~ 'Circulatory disease',
                    parent_code_num>=460 & parent_code_num<=519 ~ 'Respiratory disease',
                    parent_code_num>=520 & parent_code_num<=579 ~ 'Digestive disease',
                    parent_code_num>=580 & parent_code_num<=629 ~ 'Genitourinary System',
                    parent_code_num>=630 & parent_code_num<=679 ~ 'Pregnancy,Childbirth,Puerperium Complications',
                    parent_code_num>=680 & parent_code_num<=709 ~ 'Skindisease',
                    parent_code_num>=710 & parent_code_num<=739 ~ 'Musculoskeletal disease ',
                    parent_code_num>=740 & parent_code_num<=759 ~ 'Congenital Anomalies',
                    parent_code_num>=760 & parent_code_num<=779 ~ 'Perinatal Period conditions',
                    parent_code_num>=780 & parent_code_num<=799 ~ 'Symptoms, Signs and Ill-defined Conditions',
                    parent_code_num>=800 & parent_code_num<=999 ~ 'Injury and Poisoning'
          )
  )         
icd_reference_annotated %>% select(parent_code, parent_code_class) %>% distinct %>% 
  group_by(parent_code_class) %>% 
  summarise(start = min(parent_code), 
            end = max(parent_code)) %>% 
  arrange(start) %>% 
  mutate(`Code Interval` = paste(start, end, sep ='-')) %>% 
  select(`Code Interval`, Description = parent_code_class ) %>% 
  write_csv('ICD9_class_ref.csv')

icd9_parent_size = icd_reference_annotated %>% group_by(parent_code) %>% summarise(parent_code_size = n())


icd_code_counts_annotated <- icd_reference_annotated %>% left_join(icd_code_counts) %>% mutate(count = replace_na(count, 0)) %>% left_join(icd9_parent_size)

save(icd_code_counts_annotated, file = 'Rdata/icd_code_reference_and_counts.Rdata')

N_pt_codes_once = icd_code_counts_annotated %>% filter(count != 0) %>% nrow %>% {./nrow(icd_code_counts_annotated)} %>% {.*100} %>% round(2)
N_codes_once = icd_code_counts_annotated %>% filter(count != 0) %>% nrow


N_icd_code_less_5c <- icd_code_counts_annotated %>% filter(count < 5) %>% nrow %>% {./nrow(icd_code_counts_annotated)} %>% {.*100} %>% round(2)
```

```{r}
library(ggforce)
icd_code_counts_annotated %>% 
  mutate(k='ICD9') %>% 
ggplot(aes(x=count))+ 
  geom_density() + 
  facet_zoom(x = count < 25)
```



```{r}
icd_code_counts_annotated %>% 
  mutate(k='ICD9') %>% 
ggplot(aes(x=k, y=count))+
  geom_violin(width = .5)+
  xlab('')+
  ylim(c(0,25))+
  theme_classic() +
  theme(text = element_text(size = 15)) 

N_code_count_over_25 = sum(icd_code_counts_annotated$count > 25)
```



## SNOMED to ICD9



```{r}

icd_reference_with_snomed <- icd_reference %>% left_join(icd_to_SNOMED_mapping)
save(diag_with_admits, file = "Rdata/ICD9_diag_codes_by_patient_visit.Rdata")
save(icd_reference_with_snomed, file = "Rdata/ICD9_ref_mapped_to_snomed.Rdata")
save(icd_code_counts_annotated, file = "Rdata/ICD9_mimic3_counts.Rdata")

```

```{r}
N_1t1_icd_snomed = icd_reference_with_snomed %>% filter(type == "1-to-1") %>% pull(ICD9_CODE) %>% n_distinct
N_missing_icd_snoemd = icd_reference_with_snomed %>% filter(is.na(SNOMED_CID)) %>% select(ICD9_CODE, SNOMED_CID) %>% distinct %>% nrow
N_1tm_icd_snomed <- nrow(icd_reference) - N_1t1_icd_snomed - N_missing_icd_snoemd
N_patients_mimic <- diag_with_admits %>% select(SUBJECT_ID) %>% n_distinct
N_icd_codes_mimic <- nrow(diag_with_admits)
```


bash 
```
grep 116680003 SnomedCT_USEditionRF2_PRODUCTION_20210901T120000Z/Full/Terminology/sct2_Relationship_Full_US1000124_20210901.txt > snomed_is-a_relationships.txt

 wc -l snomed_is-a_relationships.txt
```

```{r, eval=FALSE}
#system(" python3 networkx_SNOMED_depth.py")
```

```{r}
snomed_depth <- read_csv('SNOMED_term_depth_from_networkx.csv', col_types = 'cd') %>% 
  rename(SNOMED_CID = SID)

ICD9_depth <- filter(icd_reference_with_snomed, !is.na(SNOMED_CID)) %>% 
  inner_join(snomed_depth) %>% 
  group_by(ICD9_CODE) %>% 
  summarise(avg_depth = mean(depth))

```

```{r}
icd_counts_annotated_with_depth <- icd_code_counts_annotated %>% 
  mutate(is_unused = ifelse(count == 0, 'Unused', 'Used')) %>% 
  inner_join(ICD9_depth)


N_used_icd9= icd_counts_annotated_with_depth %>% filter(is_unused == "Used") %>% nrow()
```


```{r}
snomed_vs_IC9_depth <-  bind_rows(
 snomed_depth %>% select(depth) %>% mutate(terminology = "SNOMED-CT"), 
 ICD9_depth %>% select(depth = avg_depth) %>% mutate(terminology = "ICD9CM")
) %>% 
  ggplot(aes(x=terminology, y = depth)) + 
  geom_violin() + 
  theme_minimal()

icd_counts_annotated_with_depth <- icd_counts_annotated_with_depth %>% 
  filter(!is.na(avg_depth))
plabdf <- wilcox.test(icd_counts_annotated_with_depth %>% filter(count == 0) %>% pull(avg_depth),
                      icd_counts_annotated_with_depth %>% filter(count != 0) %>% pull(avg_depth)) %>% .[['p.value']]

ggplot(icd_counts_annotated_with_depth) +
  geom_boxplot(aes(x=is_unused, y=avg_depth))  +
  xlab("")+
  ylab("average depth")+
  theme_minimal()
```


```{r}
numbers <- ls() %>% .[grepl("N_", .)]
save(list = numbers, file = 'Rdata/initial_count_data_exploration.Rdata')
```

```{r}
icd_counts_annotated_with_depth %>% filter(count == 0) %>% pull(avg_depth) %>% mean
                      icd_counts_annotated_with_depth %>% filter(count != 0) %>% pull(avg_depth) %>% mean
```
```{r}
depth_by_bin <-  icd_counts_annotated_with_depth %>% 
  filter(!is.na(avg_depth)) %>% 
  mutate(count_bin = case_when(
    count ==  0 ~ "0",
    count == 1 ~ "1", 
    count >=2 & count <5 ~ "[2,5)",
    count >=5 & count <25 ~ "[5, 25)",
    count >= 25 ~ ">=25"
  ) ) 

counts_by_bin = depth_by_bin %>% group_by(count_bin) %>% count %>% 
  mutate(count_bin = factor(count_bin, levels = c("0","1","[2,5)","[5, 25)",">=25")))

save(counts_by_bin, depth_by_bin,icd_code_counts_annotated,  icd_counts_annotated_with_depth, snomed_depth, ICD9_depth, file = 'Rdata/initial_count_data_exploration_plotdata.Rdata' )

cb <- ggplot(counts_by_bin) + 
  geom_col(aes(x=count_bin, y=n)) + theme_minimal()


db <- ggplot(depth_by_bin %>% mutate(count_bin = factor(count_bin, levels = c("0","1","[2,5)","[5, 25)",">=25")) ) )+
  geom_boxplot(aes(x = count_bin, y= avg_depth)) + 
  theme_minimal()
#cb/db

```

$ \sum_i^n e^{-W_i} * L_i + W_i $



